Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex

Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
%TODO. what we did and are going to present next.
We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.

{\bf Remark.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
for instance in Gaussian covariance estimation when $n \leq d + 1$.
Even worse with categorical variables, the MLE is always ill-defined. %TODO
Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.

\section{Insights}

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
\end{align}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix}, since
\begin{align}
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
\end{align}
From there, it is possible to show an asymptotic rate for the MLE and the MAP \textcolor{red}{D.S.: How?}
\begin{align}
\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.
We point the curious reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive review.

\subsection{Quadratic Case}
\rlp{make it multivariate, with a reference to \citep{bach2013nonstronglyconvex}?}
As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
\[ \cX=\real,\quad \nu(dx) = \exp(\half[-x^2])dx,\quad T(X)=X.\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{1}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
    &\leq 2L \norm{\m^* - \m}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
\TODO{connection with optim non-smooth rate.}

>>>>

Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.

{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
for instance in Gaussian covariance estimation when $n \leq d + 1$.
Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories. In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$. 
This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.

\section{Insights}
\label{sec:insights}

\subsection{Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now review classical asymptotic results. 
We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\alignn{
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregman_asymptote}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregman_asymptote}, one can show that MLE and MAP verify 
\begin{align}
	\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
See proof in the Appendix.
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

\subsection{Quadratic Case}
As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
\[ \cX=\real^d,\quad \nu(dx) = \exp(\half[-\|x\|^2])dx,\quad T(x)=x.\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{1}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
    &\leq 2L \norm{\m^* - \m}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
Note that none of the standard exponential families verify this assumption.
\TODO{connection with optim non-smooth rate.}
